Grove: A Convergent Collaborative Structure-Editor Calculus

### 2024-03-25 formalism.tex
feat: latest changes - uncomment sections

% TODO: point of this section is commutativity theorem

% TODO: (though as we show in \autoref{sec:Formalism} creating a vertex is implicit in edge creation)

% The formalism is rather simple, \autoref{sec:Grove By Example} shows
% that it is sufficient to handle many common situations,
% and this simplicity aids in the predictability.

% \figureArity

@@

% Let $\U$ be a set of \emph{unique identifiers} equipped with a total ordering $\leq$,
% and $\K$ the \emph{constructors} of terms in the source language,
% and $\P$ the \emph{positions} of subterms in the source language,
% and $\Sigma = \SetOf{\bot, \Plus, \Minus}$ the
%   \emph{edge states} equipped with a total ordering $\bot \sqsubset \Plus \sqsubset \Minus$ that forms a lattice over $\Sigma$.
% Now, let $\V = \U \times \K$ all possible \emph{vertices},
% and $\S = \V \times \P$ all possible \emph{sources},
% and $\E = \U \times \S \times \V$ all possible \emph{edges}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% TODO: put this in a definition box

A graph $G$ is well sorted if
for all $\e = (u, (u_1, k_1), p, (u_2, k_2))$
such that $G(\e) \neq \bot$,
we have $(p, \sort(k_2)) \in \arity(k_1)$.

@@

% In decomp section:
% - If a graph is well sorted, then there exists a grove. (i.e., exhaustiveness check for expr)
%   - want to show that expr is a total function on Exp's
% - If decomp(G) = Grove, then vertices(Grove) = vertices(edges(G))
%   - every vertex should be part of the decomposition
% - correctness of grove components:
%   - e_r: rootVertex(e_r) = v_r
%   - MP: for every expression in MP, |parents(the root vertex)| > 1
%   - NP: for every expression in NP, |parents(the root vertex)| = 0
%   - U: for every expression in U, |parents(the root vertex)| = 1 and the root vertex = min(ancestors(the root vertex))
%     - every other vertex notin U has exactly 1 parent and is not its own min-ancestor
%   - we've accounted for every edge:
%     - want to say edges(Grove) = edges(G), but we don't have a clean way to say that
%     - there is a choice from (an infinite family of graphs) recomposition(Grove) (which are all structurally equivalent to G, modulo edge ids)

% correctness covers:
% - invertibility (w/ some identity assignment)
% - parentage (things are multiparented correctly)

@@

% TODO: should graph "recomposition" have an algorithm, too?
% TODO: update this to match our current understanding
A graph can be derived from an expression in the calculus by starting with an
empty graph, selecting unique ids $u, u_k$ and constructor $k$ for the outermost
expression, and creating edge $\e = (u, v_0, p_0, (u_k, k))$. Then, for each
sub-term at some position $p \in \arity(k)$, select unique id $u'$ and
constructor $k'$ and create edge $e' = (u', v, p, v')$ with $v = (u_k, k)$, $v'
= (u', k')$. Continuing in this manner, treat each sub-term as if it were the
outermost, starting with the ones at positions in $\arity(k')$, and create edges
to the vertices representing their immediate sub-terms until only empty holes
remain. Note that empty holes are not modeled explicitly in the graph. Wherever
an empty hole appears in an expression, the corresponding graph position has no
edges originating from it.

@@

\figureTermSyntax

@@

% The first component $e_r$ is the expression derived from the root vertex $v_0$.
% \autoref{fig:Decomposition expr} defines $\expr(v)$ which determines the expression derived from $v$.
% There is one rule for each term constructor.
% For example, the expression derived from ${+}_{\id{38}} = (38, \ExpPlus)$ is $\expr'({+}_{\id{38}}, \PlusLeft) \text{+}^{\id{38}} \expr'({+}_{\id{38}}, \PlusRight)$.
% In each child position, $\expr'(v, p)$ determines the subterm for the children of $v$ at position $p$.
% An empty hole $\hole$ appears if there are no children at the given position.
% A conflict hole $\conflictHole{e_1,\ldots,e_n}$ appears if there are multiple children.
% If there is one child, the expression is determined recursively unless that child appears in $MP$ or $U$, discussed below, in which case we leave a reference, $\multiVertex{u}$ or $\cycleVertex{u}$, respectively.
% For the $\ExpLam$ case, we define $\patt(q)$ and $\type(\tau)$ similarly.

@@

% Mapping to b actions

% \[
%   \begin{array}{l}
%     \textrm{User Actions}: \grave{e} \in Act ::= e \mid \#u \mid \conflict{\grave{e}}{\grave{e}}                                                       \\
%     \s : \G \to Act \times \V^{\ast{}} \times \V^{\ast{}} \times \V^{\ast{}}                                                                                          \\
%     \s(G) = (\grave{e}, \MP(G), \CC(G), \D(G))                                                                                                         \\
%     \MP(G) = \left\{ v \in \vertexes(G) \mid \exists e_1,e_2 \in \liveEdges(G), e_1= \right.                                                           \\
%     \qquad \left. (v,p_1,v_1), e_2=(v,p_2,v_2), e_1 \ne e_2 \right\}                                                                                   \\
%     \CC(G) = \ldots                                                                                                                                    \\
%     \D(G) = \orphans(G) \setminus \left\{ v_0 \right\}                                                                                                 \\
%     \orphans(G) = \left\{ v \in \vertexes(G) \mid \forall e=(v^\prime,p,v^{\prime\prime}) \in \right.                                                  \\
%     \qquad \left. \liveEdges(G), v^{\prime\prime} \ne v \right\}                                                                                       \\
%     \vertexes(G) = \left\{ v \mid \exists e = (v^\prime, p, v^{\prime\prime}) \in \edges(G), v \in \left\{ v^\prime, v^{\prime\prime}\right\} \right\} \\
%     \edges(G) = \left\{ e \mid G(e) \ne \bot \right\}                                                                                                  \\
%     \liveEdges(G) = \left\{ e \mid G(e) = + \right\}                                                                                                   \\
%   \end{array}
% \]

>>>>

TODO: point of this section is commutativity theorem

TODO: (though as we show in \autoref{sec:Formalism} creating a vertex is implicit in edge creation)

The formalism is rather simple, \autoref{sec:Grove By Example} shows
that it is sufficient to handle many common situations,
and this simplicity aids in the predictability.

\figureArity{}

@@

Let $\U$ be a set of \emph{unique identifiers} equipped with a total ordering $\leq$,
and $\K$ the \emph{constructors} of terms in the source language,
and $\P$ the \emph{positions} of subterms in the source language,
and $\Sigma = \SetOf{\bot, \Plus, \Minus}$ the
  \emph{edge states} equipped with a total ordering $\bot \sqsubset \Plus \sqsubset \Minus$ that forms a lattice over $\Sigma$.
Now, let $\V = \U \times \K$ all possible \emph{vertices},
and $\S = \V \times \P$ all possible \emph{sources},
and $\E = \U \times \S \times \V$ all possible \emph{edges}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% TODO: put this in a definition box
A graph $G$ is well sorted if
for all $\e = (u, (u_1, k_1), p, (u_2, k_2))$
such that $G(\e) \neq \bot$,
we have $(p, \sort (k_2)) \in \arity k_1$.

@@

In decomp section:
- If a graph is well sorted, then there exists a grove. (i.e., exhaustiveness check for expr)
  - want to show that expr is a total function on Exps
- If decomp(G) = Grove, then vertices(Grove) = vertices(edges(G))
  - every vertex should be part of the decomposition
- correctness of grove components:
  - $e_r$: rootVertex($e_r) = v_r$
  - MP: for every expression in MP, |parents(the root vertex)| > 1
  - NP: for every expression in NP, |parents(the root vertex)| = 0
  - U: for every expression in U, |parents(the root vertex)| = 1 and the root vertex = min(ancestors(the root vertex))
    - every other vertex notin U has exactly 1 parent and is not its own min-ancestor
  - we've accounted for every edge:
    - want to say edges(Grove) = edges(G), but we don't have a clean way to say that
    - there is a choice from (an infinite family of graphs) recomposition(Grove) (which are all structurally equivalent to G, modulo edge ids)

correctness covers:
- invertibility (w/ some identity assignment)
- parentage (things are multiparented correctly)

@@

% TODO: should graph "recomposition" have an algorithm, too?
% TODO: update this to match our current understanding
A graph can be derived from an expression in the calculus by starting with an
empty graph, selecting unique ids $u, u_k$ and constructor $k$ for the outermost
expression, and creating edge $\e = (u, v_0, p_0, (u_k, k))$. Then, for each
sub-term at some position $p \in \arity k$, select unique id $u'$ and
constructor $k'$ and create edge $e' = (u', v, p, v')$ with $v = (u_k, k)$, $v'
= (u', k')$. Continuing in this manner, treat each sub-term as if it were the
outermost, starting with the ones at positions in $\arity k'$, and create edges
to the vertices representing their immediate sub-terms until only empty holes
remain. Note that empty holes are not modeled explicitly in the graph. Wherever
an empty hole appears in an expression, the corresponding graph position has no
edges originating from it.

@@

\figureTermSyntax{}

@@

The first component $e_r$ is the expression derived from the root vertex $v_0$.
\autoref{fig:Decomposition expr} defines $\expr(v)$ which determines the expression derived from $v$.
There is one rule for each term constructor.
For example, the expression derived from ${+}_{\id{38}} = (38, \ExpPlus)$ is $\expr'({+}_{\id{38}}, \PlusLeft) \text{+}^{\id{38}} \expr'({+}_{\id{38}}, \PlusRight)$.
In each child position, $\expr'(v, p)$ determines the subterm for the children of $v$ at position $p$.
An empty hole $\hole$ appears if there are no children at the given position.
A conflict hole $\conflictHole{e_1,\ldots,e_n}$ appears if there are multiple children.
If there is one child, the expression is determined recursively unless that child appears in $MP$ or $U$, discussed below, in which case we leave a reference, $\multiVertex{u}$ or $\cycleVertex{u}$, respectively.
For the $\ExpLam$ case, we define $\patt(q)$ and $\type(\tau)$ similarly.

@@

Mapping to b actions

\[
  \begin{array}{l}
    \textrm{User Actions}: \grave{e} \in Act ::= e \mid \#u \mid \conflict{\grave{e}}{\grave{e}}                                                       \\
    \s : \G \to Act \times \V^{\ast{}} \times \V^{\ast{}} \times \V^{\ast{}}                                                                                          \\
    \s(G) = (\grave{e}, \MP(G), \CC(G), \D(G))                                                                                                         \\
    \MP(G) = \left\{ v \in \vertexes(G) \mid \exists e_1,e_2 \in \liveEdges(G), e_1= \right.                                                           \\
    \qquad \left. (v,p_1,v_1), e_2=(v,p_2,v_2), e_1 \ne e_2 \right\}                                                                                   \\
    \CC(G) = \ldots                                                                                                                                    \\
    \D(G) = \orphans(G) \setminus \left\{ v_0 \right\}                                                                                                 \\
    \orphans(G) = \left\{ v \in \vertexes(G) \mid \forall e=(v^\prime,p,v^{\prime\prime}) \in \right.                                                  \\
    \qquad \left. \liveEdges(G), v^{\prime\prime} \ne v \right\}                                                                                       \\
    \vertexes(G) = \left\{ v \mid \exists e = (v^\prime, p, v^{\prime\prime}) \in \edges(G), v \in \left\{ v^\prime, v^{\prime\prime}\right\} \right\} \\
    \edges(G) = \left\{ e \mid G(e) \ne \bot \right\}                                                                                                  \\
    \liveEdges(G) = \left\{ e \mid G(e) = + \right\}                                                                                                   \\
  \end{array}
\]
